---
title: "Analysis of predicted protein levels in tumour samples"
author: A. Hanson
date: "`r format(Sys.time(), '%d %B %Y')`"
format: 
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: true
---

```{r}
#| label: setup
#| include: false

#library(remotes)
#install_github("perishky/meffonym")

library(gtsummary)
library(gt)
library(cowplot)
library(ggrepel)
library(dplyr)
library(ggplot2)
library(here)
library(meffonym) 

# Read environment file
readRenviron(here("config.env"))

datadir <- Sys.getenv("datadir")
resultsdir <- Sys.getenv("resultsdir")

#knitr::opts_chunk$set(fig.path = "figures/")
```

### Introduction

Here we are analysing DNA methylation-predicted protein abundances in head and neck squamous cell carcinoma (HNSC) samples and normal tissue from [The Cancer Genome Atlas (TGCA)](https://gdc.cancer.gov/about-data/publications/pancanatlas). Predicted levels of 109 proteins have been derived using DNA methylation data with the [`meffonym`](https://github.com/perishky/meffonym) R package. Paired clinical outcome data is available through the PanCancer Atlas.

Cohort characteristics are provided in @tbl-summary.

```{r}
#| label: load data
#| include: FALSE

filename <- file.path(resultsdir, "combined-clin-pred-proteins.txt")
# Load data
data <- read.table(filename, header = T, sep = "\t", stringsAsFactors = F)

# Protein names
protein.names <- subset(meffonym::meffonym.models(full = T),
    grepl("^episcores", filename))$name
protein.names <- make.names(protein.names)
```

```{r}
#| label: tbl-summary

# Summary table
t1 <- tbl_summary(select(data, 
                         c(female, 
                           histology, 
                           age.at.diagnosis,
                           ethnicity,
                           race,
                           stage,
                           tumor.or.normal)), 
                  missing = "no",
                  label = list(
                    female = "Female",
                    histology = "Histology",
                    age.at.diagnosis = "Age at diagnosis",
                    ethnicity = "Ethnicity",
                    race = "Race",
                    stage = "Stage"), 
                  by = tumor.or.normal) |> 
  modify_caption("**Table 1: Cohort Characteristics**") |>
  bold_labels() |>
  as_gt() |>
  gt::cols_width(
    label ~ gt::px(250),    # Set first column to 250px
    everything() ~ gt::px(150))  # Set other columns to 150px

t1
```

### Analysis

#### Protien levels vs tissue type

Is proteins abundance predictive of tissue type (tumour vs normal)?\
Model: `glm(tissue ~ predicted protein, family = binomial)`

```{r}
#| label: analysis-1

# Define glm formule with pred.proteins as predictors of tumor.or.normal
formulae <- sapply(protein.names, function(i) {
  reformulate(i, response = "tumor")
}, simplify = F)

# Run glms
fit <- sapply(formulae, function(i) {
  glm(i, data = data, family = binomial())
}, simplify = F)

# Model summaries
fit.summary <- sapply(fit, function(i) {
  out <- summary(i)$coefficients
  out[, "Estimate"] <- out[, "Estimate"]
  out
}, simplify = F)

# Make results table
fit.coefs <- sapply(fit.summary, function(i) {
  i[2, c("Estimate", "Pr(>|z|)")]
}, simplify = F)

fit.coefs <- {
  x <- do.call(rbind, fit.coefs)
  data.frame(
    pred.protein = rownames(x),
    coef = x[, "Estimate"],
    p.value = x[, "Pr(>|z|)"]
  )
}

bonferroni <- -log10(0.05 / length(fit))
```

There were `r sum(-log10(fit.coefs$p.value) > bonferroni)` proteins that associated with tissue type following correction for multiple testing, when assessed using binomial logistic regression (@fig-analysis-1).

```{r}
#| label: fig-analysis-1
#| fig-cap: "Predicted proteins significantly associated with tissue type following multiple testing correction (a) and volcano plot of model effect estimates (b). Bonferroni threshold for significance is indicated by the red dashed line, points in b are scaled by absolute effect size."
#| fig-width: 14
#| fig-height: 5
#| warning: FALSE

p1.1 <- fit.coefs |>
  ggplot(aes(x = pred.protein, y = -log10(p.value))) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed",
    colour = "red",
  ) +
  geom_point(size = 3, 
             alpha = 0.6) +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = pred.protein, y = -log10(p.value), label = pred.protein),
    colour = "blue",
    size = 3,
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7)) +
  labs(y = "-log10(P)", x = "Metylation predicted protein")

# Volcano plot
p1.2 <- fit.coefs |>
  ggplot(aes(x = coef, y = -log10(p.value))) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed",
    colour = "red"
  ) +
  geom_point(aes(size = abs(coef)), 
             alpha = 0.6, 
             show.legend = FALSE) +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = coef, y = -log10(p.value), label = pred.protein),
    colour = "blue",
    size = 3,
  ) +
  theme_minimal() +
  labs(y = "-log10(P)", x = "Coefficient")

cowplot::plot_grid(plotlist = list(p1.1,p1.2), nrow = 1, labels = "auto", rel_widths = c(2,1))

```

#### Protien levels vs progression free interval (pfi)

Is proteins abundance predictive of tumour free progression (binary 0/1)?\
Model: `glm(pfi ~ predicted protein, family = binomial)`

```{r}
#| label: analysis-2

# Restrict analysis to measurments taken from tumour samples 
tumor.data <- subset(data, tumor == 1)

# Define glm formule with pred.proteins as predictors of pfi (coded 0 or 1)
formulae <- sapply(protein.names, function(i) {
  reformulate(i, response = "pfi")
}, simplify = F)

# Run GLMS
fit <- sapply(formulae, function(i) {
  glm(i, data = tumor.data, family = binomial())
}, simplify = F)

# Model summaries
fit.summary <- sapply(fit, function(i) {
  out <- summary(i)$coefficients
  out[, "Estimate"] <- out[, "Estimate"]
  out
}, simplify = F)

# Results table
fit.coefs <- sapply(fit.summary, function(i) {
  i[2, c("Estimate", "Pr(>|z|)")]
}, simplify = F)

fit.coefs <- {
  x <- do.call(rbind, fit.coefs)
  data.frame(
    pred.protein = rownames(x),
    coef = x[, "Estimate"],
    p.value = x[, "Pr(>|z|)"]
  )
}
```

There was `r sum(-log10(fit.coefs$p.value) > bonferroni)` protein that associated with progression free interval, following correction for multiple testing, when assessed using binomial logistic regression (@fig-analysis-2).

```{r}
#| label: fig-analysis-2
#| fig-cap: "Predicted proteins significantly associated with progression free interval following multiple testing correction (a) and volcano plot of model effect estimates (b). Bonferroni threshold for significance is indicated by the red dashed line, points in b are scaled by absolute effect size."
#| fig-width: 14
#| fig-height: 5
#| warning: FALSE

p2.1 <- fit.coefs |>
  ggplot(aes(x = pred.protein, y = -log10(p.value))) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed",
    colour = "red",
  ) +
  geom_point(size = 3, 
             alpha = 0.6) +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = pred.protein, y = -log10(p.value), label = pred.protein),
    colour = "blue",
    size = 5,
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7)) +
  labs(y = "-log10(P)", x = "Metylation predicted protein")

# Volcano plot
p2.2 <- fit.coefs |>
  ggplot(aes(x = coef, y = -log10(p.value))) +
  geom_hline(
    yintercept = bonferroni,
    linetype = "dashed",
    colour = "red"
  ) +
  geom_point(aes(size = abs(coef)), 
             alpha = 0.6, 
             show.legend = FALSE) +
  geom_text_repel(
    data = fit.coefs[which((-log10(fit.coefs$p.value) > bonferroni)), ],
    aes(x = coef, y = -log10(p.value), label = pred.protein),
    colour = "blue",
    size = 5,
  ) +
  theme_minimal() +
  labs(y = "-log10(P)", x = "Coefficient")

cowplot::plot_grid(plotlist = list(p2.1,p2.2), nrow = 1, labels = "auto", rel_widths = c(2,1))
```
